Machine Learning Approaches for Accurate Diamond Price Prediction\\ \thanks{Final Project Report ECS-171: \textbf{Group 21}}

### 2024-06-03 paper/main.tex
neural network example

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{natbib}
\usepackage{float}
\usepackage{xcolor}
\bibliographystyle{plainnat}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

\author{
\IEEEauthorblockN{Aditya Mittal}
\IEEEauthorblockA{\textit{Department of Statistics} \\
\textit{University of California, Davis}\\
Davis, United States \\
adimittal@ucdavis.edu}
\and
\IEEEauthorblockN{Andrew Yeow}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University of California, Davis}\\
Davis, United States \\
XXX}
\and
\IEEEauthorblockN{Yifan Cui}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University of California, Davis}\\
Davis, United States \\
XXX}
\and
\IEEEauthorblockN{Nandini Koladi}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University of California, Davis}\\
Davis, United States \\
XXX}
\and
\IEEEauthorblockN{Eshan Toshniwal}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University of California, Davis}\\
Davis, United States \\
XXX}
}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.745\linewidth]{Figure_1.png}
    \caption{Distribution Plot of Continuous Features. Identify trends in our dataset such as skewness, location of outliers, mean, median, scale, etc.}
    \label{fig:image_label}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\linewidth]{Figure_4.png} % Replace "example-image" with the filename of your image
    \caption{Pair-wise Scatterplots for Each Feature vs. Diamond Price}
    \label{fig:image_label}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\linewidth]{Figure_3.png} % Replace "example-image" with the filename of your image
    \caption{Correlation Heat-Map for Continuous Features vs Diamond Price}
    \label{fig:image_label}
\end{figure}

@@

@@

In our exploration of predicting diamond prices, we employ artificial neural networks (ANNs). Our ANN architecture consists of a single hidden layer with ReLU activation, followed by an output layer also utilizing ReLU activation, as our task involves regression. To optimize the performance of our ANN, we conduct a grid search for hyperparameter tuning, aiming to select the most suitable parameters for our model. We explore the following hyperparameters during grid search with a 5-fold cross-validation procedure. We explore the following set of hyperparameters:

@@

\begin{table}[H]
\centering
\caption{Optimal Hyperparameter for Random Forests}
\begin{tabular}{|c|c|}
\hline
Hyperparameter & Optimal Value \\
\hline
Learning Rate & XXX \\
\hline
Epochs & XXX \\
\hline
Batch Size & XXX \\
\hline
Optimizer & XXX \\
\hline
Dropout Rate & XXX \\
\hline
Number of Neurons in Hidden Layer & XXX \\
\hline
Number of Hidden Layers & XXX \\
\hline
\end{tabular}
\label{tab:mytable}
\end{table}

The grid search method results in the architecture of our neural network model. Specifically, [...] Here, we present the training versus testing loss (MSE) across epochs during the training of our neural network model, along with its final performance on the testing data.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\linewidth]{train vs test.png} % Replace "example-image" with the filename of your image
    \caption{Training vs Testing MSE Across Epoch}
    \label{fig:boxplots}
\end{figure}

@@

\begin{table}[H]
\centering
\caption{Neural Network Performance on Testing Data}
\begin{tabular}{|c|c|c|c|}
\hline
Method & MSE & MAE & $R^2$ \\
\hline
Neural Network & XXX & XXX & XXX \\
\hline
\end{tabular}
\label{tab:mytable}
\end{table}

@@

\begin{table}[H]
\centering
\caption{Comparison of Algorithms}
\begin{tabular}{|c|c|c|c|c|}
\hline
Method & MSE & MAE & $R^2$ & BIC \\
\hline
Linear Regression & 1,110,908.273 & 718.843 & 0.898 & XXX \\
\hline
Polynomial Regression & 459,504.630 & 381.455 &  0.958 & XXX \\
\hline
Random Forests & 228,347.747 & 247.141 & 0.9803 & XXX \\
\hline
XGBoost & 195,937.71 & 233.941 & 0.9831 & XXX \\
\hline
Neural Networks & XXX & XXX & XXX & XXX \\
\hline
\end{tabular}
\label{tab:mytable}
\end{table}

[ADD ANALYSIS]

@@

\section{Project RoadMap}

Our project involves a regression-supervised learning task in the context of predicting the price of a diamond based on several different measures. Our dataset contains 53,940 observations and 9 columns (8 predictors and 1 response). The output variable “price” is continuous and has a range between $326 to $18,823. With this, our goal for this project is to create a machine-learning model using deep neural networks that can predict diamond prices as accurately as possible. The practical application of this project is quite significant, as an effective model can assist researchers and industry professionals in determining the value of diamonds in the market. 
As a part of developing our machine learning pipeline, we will first begin with exploratory data analysis. We do not have any missing values present, so we don’t have to worry about any data imputation for this specific project. First, we will begin by plotting the distribution of our features to understand the spread of our data and account for potential skewness. Since outliers can have significant effects on the overall accuracy of machine learning models, we will look for them in our dataset and remove them. We will use various outlier detection techniques such as one-class SVM and through the IQR computation; we're still planning the specific details as to how we will approach this step. Furthermore, since we have 3 categorical variables, we will use one-hot encoding and label encoding (in the case of an ordinal column) to convert them into numeric columns. Ultimately, our goal with EDA is to clean the dataset and understand the relationship between the predictors and the response variable. In terms of timeline, we plan to dedicate the EDA portion of our dataset this week/early next week (~ 1/1.5 weeks) before we begin training the model. We aim to have this portion of the project completed before the quarter midpoint and begin the process of training the model as well.

@@

\newpage
\begin{thebibliography}{00}
\bibitem{kino} Kino, S., Hsu, Y.-T., Shiba, K., Chien, Y.-S., Mita, C., Kawachi, I., and Daoud, A. (2021). A scoping review on the use of machine learning in research on social determinants of health: Trends and research prospects. SSM - Population Health, 15, 100836.
\bibitem{sarker} Sarker, I.H. Machine Learning: Algorithms, Real-World Applications and Research Directions. SN COMPUT. SCI. 2, 160 (2021). https://doi.org/10.1007/s42979-021-00592-x
\bibitem{Alsuraihi} Alsuraihi, W., Al-hazmi, E., Bawazeer, K., \& AlGhamdi, H. Machine learning algorithms for diamond price prediction. In Proceedings of the 2020 2nd International Conference on Image, Video and Signal Processing, 150–154 (2020).
\bibitem{Mamonov} Mamonov, S. \& Triantoro, T. Subjectivity of diamond prices in online retail: Insights from a data mining study. J. Theor. Appl. Electron. Commer. Res. 13(2), 15–28 (2018).
\bibitem{Pandey} Pandey, A. C., Misra, S., \& Saxena, M. Gold and diamond price prediction using enhanced ensemble learning. In 2019 Twelfth International Conference on Contemporary Computing (IC3), 1–4 (IEEE, 2019).
\bibitem{Scott} Scott, F. \& Yelowitz, A. Pricing anomalies in the market for diamonds: Evidence of conformist behavior. Econ. Inq. 48(2), 353–368 (2010).
\bibitem{Kutner}Michael Kutner, Christopher Nachtsheim, John Neter, and William Li. Applied Linear Statistical Models. 1974.
\bibitem{Seber}Seber, G. A. F., \& Lee, A. J. (2012). Linear Regression Analysis. Hoboken, NJ: John Wiley \& Sons.
\bibitem{Wright} Wright, John T. "Linear Regression Analysis." British Medical Journal, vol. 310, no. 6977, 1995, pp. 1120-1124. BMJ Group
\bibitem{Weisberg} Weisberg, S. (2005). Applied Linear Regression. Wiley.
\bibitem{Breiman} Breiman, L. (2001). Random forests. Machine Learning, 45(1), 5-32.
\bibitem{Chen} Chen, T., \& Guestrin, C. (2016). XGBoost: A Scalable Tree Boosting System. In Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining (pp. 785-794).
\bibitem{LeCun} LeCun, Y., Bengio, Y., \& Hinton, G. (2015). Deep learning. Nature, 521(7553), 436-444.
\bibitem{Rumelhart} Rumelhart, D. E., Hinton, G. E., \& Williams, R. J. (1986). Learning representations by back-propagating errors. Nature, 323(6088), 533-536.
\bibitem{Bergstra} Bergstra, J., \& Bengio, Y. (2012). Random search for hyper-parameter optimization. Journal of Machine Learning Research, 13(Feb), 281-305.
\bibitem{us} U.S. jewelry market size and share: Industry Report, 2030. U.S. Jewelry Market Size And Share | Industry Report, 2030. (n.d.). https://www.grandviewresearch.com/industry-analysis/us-jewelry-market-report
\bibitem{diamond} Diamond Jewelry Market Size \& Share Analysis Report, 2030. (n.d.). https://www.grandviewresearch.com/industry-analysis/diamond-jewelry-market-report
\bibitem{kaggle} Agrawal, S. (2017a, May 25). Diamonds. Kaggle. https://www.kaggle.com/datasets/shivam2503/diamonds
\bibitem{Bottou} Bottou, L. (2010). Large-Scale Machine Learning with Stochastic Gradient Descent. In Proceedings of COMPSTAT'2010, Springer. DOI: 10.1007/978-3-7908-2604-3\_16
\bibitem{Kingma} Kingma, D. P., \& Ba, J. (2015). Adam: A Method for Stochastic Optimization. In International Conference on Learning Representations (ICLR). URL: https://arxiv.org/abs/1412.6980
\bibitem{Tieleman} Tieleman, T., \& Hinton, G. (2012). Lecture 6.5 - RMSProp: Divide the gradient by a running average of its recent magnitude. COURSERA: Neural Networks for Machine Learning. URL: https://www.cs.toronto.edu/~tijmen/csc321/slides/lecture\_slides\_lec6.pdf
\end{thebibliography}
\vspace{12pt}

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
%\usepackage{subcaption}
\usepackage{natbib}
\usepackage{float}
\usepackage{xcolor}
\bibliographystyle{plainnat}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

\author{
\IEEEauthorblockN{Aditya Mittal}
\IEEEauthorblockA{\textit{Department of Statistics} \\
\textit{University of California, Davis}\\
Davis, United States \\
adimittal@ucdavis.edu}
\and
\IEEEauthorblockN{Andrew Yeow}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University of California, Davis}\\
Davis, United States \\
XXX}
\and
\IEEEauthorblockN{Yifan Cui}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University of California, Davis}\\
Davis, United States \\
harcui@ucdavis.edu}
\and
\IEEEauthorblockN{Nandini Koladi}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University of California, Davis}\\
Davis, United States \\
XXX}
\and
\IEEEauthorblockN{Eshan Toshniwal}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University of California, Davis}\\
Davis, United States \\
XXX}
}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{Figure_1.png}
    \caption{Distribution Plot of Continuous Features. Identify trends in our dataset such as skewness, location of outliers, mean, median, scale, etc.}
    \label{fig:image_label}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{Figure_4.png} % Replace "example-image" with the filename of your image
    \caption{Pair-wise Scatterplots for Each Feature vs. Diamond Price}
    \label{fig:image_label}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{Figure_3.png} % Replace "example-image" with the filename of your image
    \caption{Correlation Heat-Map for Continuous Features vs Diamond Price}
    \label{fig:image_label}
\end{figure}

@@

\subsection{Regression}

@@

In our exploration of predicting diamond prices, we employ neural networks. Our neural network architecture consists of a single hidden layer with ReLU activation, followed by an output layer also utilizing ReLU activation, as our task involves regression. To optimize the performance of our model, we conduct a grid search for hyperparameter tuning, aiming to select the most suitable parameters for our model. We explore the following hyperparameters during grid search with a 5-fold cross-validation procedure. We explore the following set of hyperparameters:

@@

\begin{table}[H]
\centering
\caption{Optimal Hyperparameter for Random Forests}
\begin{tabular}{|c|c|}
\hline
Hyperparameter & Optimal Value \\
\hline
Learning Rate & 0.1 \\
\hline
Epochs & 250 \\
\hline
Batch Size & 256 \\
\hline
Optimizer & RMSProp \\
\hline
Dropout Rate & None \\
\hline
Number of Neurons in Hidden Layer & 64 \\
\hline
\end{tabular}
\label{tab:mytable}
\end{table}

The grid search method results in the architecture of our neural network model. Specifically, the neural network architecture seems to consist of a single hidden layer with 64 neurons, employing a learning rate of 0.1, RMSProp optimizer, and no dropout regularization. It underwent training for 250 epochs with a batch size of 256. 
Here, we present the training versus testing loss (MSE) across epochs during the training of our neural network model, along with its final performance on the testing data.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\linewidth]{1000.png} % Replace "example-image" with the filename of your image
    \caption{MAE vs Degree of Polynomial Regression model}
    \label{fig:boxplots}
\end{figure}

From our results of training the neural network across 1000 epochs, we observe that the model effectively learns the weights. Specifically, both the training and testing errors significantly reduce during the initial stages of training. However, around 500 epochs, our neural network model begins to overfit the data: the testing error starts to increase significantly while the training error continues to decrease. This indicates that we can use 250 epochs as our stopping point for neural network training to avoid overfitting.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\linewidth]{250.png} % Replace "example-image" with the filename of your image
    \caption{MAE vs Degree of Polynomial Regression model}
    \label{fig:boxplots}
\end{figure}

@@

\begin{table}[H]
\centering
\caption{Neural Network Performance on Testing Data}
\begin{tabular}{|c|c|c|c|}
\hline
Method & MSE & MAE & $R^2$ \\
\hline
Neural Network & 382,237.13 & 272.47 & 0.9652  \\
\hline
\end{tabular}
\label{tab:mytable}
\end{table}

@@

\begin{table}[H]
\centering
\caption{Comparison of Algorithms}
\begin{tabular}{|c|c|c|c|c|}
\hline
Method & MSE & MAE & $R^2$ \\
\hline
Linear Regression & 1,110,908.273 & 718.843 & 0.898 \\
\hline
Polynomial Regression & 459,504.630 & 381.455 &  0.958 \\
\hline
Random Forests & 228,347.747 & 247.141 & 0.9803  \\
\hline
XGBoost & 195,937.71 & 233.941 & 0.9831 \\
\hline
Neural Networks & 382,237.13 & 272.47 & 0.9652 \\
\hline
\end{tabular}
\label{tab:mytable}
\end{table}

The comparison of algorithms for diamond pricing reveals that XGBoost performs the best, achieving the lowest MSE (195,937.71) and MAE (233.941), along with the highest $R^2$ (0.9831), indicating it provides the most accurate price predictions and best explains the variance in diamond prices. Random Forests also demonstrate strong performance, closely following XGBoost. Neural Networks and Polynomial Regression are effective but exhibit higher error rates and slightly lower $R^2$ values, making them less precise. Linear Regression, though straightforward and easy to interpret, performs the worst with the highest MSE and MAE, and the lowest $R^2$, suggesting it is the least reliable for predicting diamond prices. Therefore, ensemble methods like XGBoost and Random Forests are the most effective models for accurate diamond pricing, while Linear Regression is the least suitable.

@@

\section{Project Road-Map}

Our project involves a regression-supervised learning task in the context of predicting the price of a diamond based on several different measures. Our dataset contains 53,940 observations and 9 columns (8 predictors and 1 response). The output variable “price” is continuous and has a range between \$326 to \$18,823. With this, our goal for this project is to create a machine-learning model using deep neural networks that can predict diamond prices as accurately as possible. The practical application of this project is quite significant, as an effective model can assist researchers and industry professionals in determining the value of diamonds in the market. 
As a part of developing our machine learning pipeline, we will first begin with exploratory data analysis. We do not have any missing values present, so we don’t have to worry about any data imputation for this specific project. First, we will begin by plotting the distribution of our features to understand the spread of our data and account for potential skewness. Since outliers can have significant effects on the overall accuracy of machine learning models, we will look for them in our dataset and remove them. We will use various outlier detection techniques such as one-class SVM and through the IQR computation; we're still planning the specific details as to how we will approach this step. Furthermore, since we have 3 categorical variables, we will use one-hot encoding and label encoding (in the case of an ordinal column) to convert them into numeric columns. Ultimately, our goal with EDA is to clean the dataset and understand the relationship between the predictors and the response variable. In terms of timeline, we plan to dedicate the EDA portion of our dataset this week/early next week (~ 1/1.5 weeks) before we begin training the model. We aim to have this portion of the project completed before the quarter midpoint and begin the process of training the model as well.

@@

\newpage
\begin{thebibliography}{00}
\bibitem{kino} Kino, S., Hsu, Y.-T., Shiba, K., Chien, Y.-S., Mita, C., Kawachi, I., and Daoud, A. (2021). A scoping review on the use of machine learning in research on social determinants of health: Trends and research prospects. SSM - Population Health, 15, 100836.
\bibitem{sarker} Sarker, I.H. Machine Learning: Algorithms, Real-World Applications and Research Directions. SN COMPUT. SCI. 2, 160 (2021). https://doi.org/10.1007/s42979-021-00592-x
\bibitem{Alsuraihi} Alsuraihi, W., Al-hazmi, E., Bawazeer, K., \& AlGhamdi, H. Machine learning algorithms for diamond price prediction. In Proceedings of the 2020 2nd International Conference on Image, Video and Signal Processing, 150–154 (2020).
\bibitem{Mamonov} Mamonov, S. \& Triantoro, T. Subjectivity of diamond prices in online retail: Insights from a data mining study. J. Theor. Appl. Electron. Commer. Res. 13(2), 15–28 (2018).
\bibitem{Pandey} Pandey, A. C., Misra, S., \& Saxena, M. Gold and diamond price prediction using enhanced ensemble learning. In 2019 Twelfth International Conference on Contemporary Computing (IC3), 1–4 (IEEE, 2019).
\bibitem{Scott} Scott, F. \& Yelowitz, A. Pricing anomalies in the market for diamonds: Evidence of conformist behavior. Econ. Inq. 48(2), 353–368 (2010).
\bibitem{Kutner}Michael Kutner, Christopher Nachtsheim, John Neter, and William Li. Applied Linear Statistical Models. 1974.
\bibitem{Seber}Seber, G. A. F., \& Lee, A. J. (2012). Linear Regression Analysis. Hoboken, NJ: John Wiley \& Sons.
\bibitem{Wright} Wright, John T. "Linear Regression Analysis." British Medical Journal, vol. 310, no. 6977, 1995, pp. 1120-1124. BMJ Group
\bibitem{Weisberg} Weisberg, S. (2005). Applied Linear Regression. Wiley.
\bibitem{Breiman} Breiman, L. (2001). Random forests. Machine Learning, 45(1), 5-32.
\bibitem{Chen} Chen, T., \& Guestrin, C. (2016). XGBoost: A Scalable Tree Boosting System. In Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining (pp. 785-794).
\bibitem{LeCun} LeCun, Y., Bengio, Y., \& Hinton, G. (2015). Deep learning. Nature, 521(7553), 436-444.
\bibitem{Rumelhart} Rumelhart, D. E., Hinton, G. E., \& Williams, R. J. (1986). Learning representations by back-propagating errors. Nature, 323(6088), 533-536.
\bibitem{Bergstra} Bergstra, J., \& Bengio, Y. (2012). Random search for hyper-parameter optimization. Journal of Machine Learning Research, 13(Feb), 281-305.
\bibitem{us} U.S. jewelry market size and share: Industry Report, 2030. U.S. Jewelry Market Size And Share | Industry Report, 2030. (n.d.). https://www.grandviewresearch.com/industry-analysis/us-jewelry-market-report
\bibitem{diamond} Diamond Jewelry Market Size \& Share Analysis Report, 2030. (n.d.). https://www.grandviewresearch.com/industry-analysis/diamond-jewelry-market-report
\bibitem{kaggle} Agrawal, S. (2017a, May 25). Diamonds. Kaggle. https://www.kaggle.com/datasets/shivam2503/diamonds
\bibitem{Bottou} Bottou, L. (2010). Large-Scale Machine Learning with Stochastic Gradient Descent. In Proceedings of COMPSTAT'2010, Springer. DOI: 10.1007/978-3-7908-2604-3\_16
\bibitem{Kingma} Kingma, D. P., \& Ba, J. (2015). Adam: A Method for Stochastic Optimization. In International Conference on Learning Representations (ICLR). URL: https://arxiv.org/abs/1412.6980
\bibitem{Tieleman} Tieleman, T., \& Hinton, G. (2012). Lecture 6.5 - RMSProp: Divide the gradient by a running average of its recent magnitude. COURSERA: Neural Networks for Machine Learning. URL: https://www.cs.toronto.edu/~tijmen/csc321/slides/lecture\_slides\_lec6.pdf

\end{thebibliography}
\vspace{12pt}
